A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-12 pkg/colorspace/paper/Toolbox/article.Rmd
some feedback from RS, and started rewriting intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@483 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

  To aid selection and application of these palettes the package provides scales
  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
  exploration, visualizations of palette properties,
  accompanying manipulation utilities (like desaturation and lighten/darken), and
  emulation of color vision deficiencies.
keywords:
  # at least one keyword, comma-separated and in sentence case
  formatted: [color, palette, HCL, RGB, hue, color vision deficiency, "\\proglang{R}"]
  plain:     [color, palette, HCL, RGB, hue, color vision deficiency, R]
preamble: >
  \usepackage{thumbpdf,lmodern}
  \usepackage{booktabs,amsmath,longtable,supertabular,framed}
  \newcommand{\tex}[1]{#1}
output: rticles::jss_article
bibliography: ../../vignettes/color.bib
---

@@

The HCL space (= polar coordinates in CIELUV) is particularly useful for
specifying individual colors and color palettes as its three axes match those
of the human visual system very well: Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.

@@

To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
and sequential `"Purples 3"` palettes that were also employed above
(see Figure\ \ref{fig:iris-diamonds}). For the first example, semi-transparent
shaded densities of the sepal length from the `iris` data are shown, grouped by species
 (see the left panel of Figure\ \ref{fig:iris-ggplot}).

@@

And for the second example the sequential palette is used to code the cut levels
in a scatter of price by carat in the `diamonds` data (or rather a small subsample
thereof, see the right panel of Figure\ \ref{fig:iris-ggplot}). The scale function first
generates six colors but then drops the first color because the light gray is too light
here. (Alternatively, the chroma and luminance parameters could also be tweaked.)

@@

In Figure\ \ref{fig:applots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
application for a sequential palette. The elevation of the volcano is brought out clearly,
focusing with the dark colors on the higher elevations.
The other two displays show that hue is constant in the palette while luminance and chroma vary.
Luminance increases monotonically from dark to light (as required for a proper sequential palette).
Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
when compared to a monotonic chroma trajectory.

@@

For distinguishing colors in the middle, two strategies can be employed:
(a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
hue `h1` and ending hue `h2`). More details are provided in the next section.
(b) Instead of a decreasing chroma a triangular chroma trajectory can be employed
from `c1` over `cmax` to `c2` (or equivalently a vector `c` of
length 3). This yields high-chroma colors in the middle of the palette that are
more easily distinguished from the dark and light extremes. See `"Blues 3"`, `"Purples 3"`,
`"Reds 3"`, and `"Greens 3"` in Figure\ \ref{fig:hcl-palettes-sequentials}.

@@

* Single-hue: In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs.\ light gray).
* Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
* Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

@@

- Hue is drawn in red and coordinates are indicated on the axis on the right
  with range $[-360, 360]$.
- Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
  is used unless the palette necessitates higher chroma values.
- Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].

>>>>

  To aid selection and application of these palettes the package provides scales
  for use with \pkg{ggplot2}, \pkg{shiny} (and \pkg{tcltk}) apps for interactive
  exploration, visualizations of palette properties,
  accompanying manipulation utilities (like desaturation and lighten/darken), and
  emulation of color vision deficiencies.
keywords:
  # at least one keyword, comma-separated and in sentence case
  formatted: [color, palette, HCL, RGB, hue, color vision deficiency, "\\proglang{R}"]
  plain:     [color, palette, HCL, RGB, hue, color vision deficiency, R]
preamble: >
  \usepackage{thumbpdf,lmodern}
  \usepackage{booktabs,amsmath,longtable,supertabular,framed}
  \shortcites{color:colorspace,color:ggplot2pkg}
  \newcommand{\tex}[1]{#1}
output: rticles::jss_article
bibliography: ../../vignettes/color.bib
---

@@

Color is an integral and omnipresent element of many statistical graphics and data
visualizations. Therefore, colors should be carefully chosen to support the information
displayed [@color:Tufte:1990, @color:Brewer:1999, @color:Ware:2004, @color:Wilkinson:2005, @color:Wilke:2019].
However, until relatively recently many software packages have been using color
palettes derived from simple RGB (red-green-blue)
color combinations such as the RGB "rainbow" (or "jet") color palette
with poor perceptual properties. See @color:Hawkins+McNeall+Stephenson:2014
and @color:Stauffer+Mayr+Dabernig:2015 and the references therein for an overview.

To address these problems, many improved color palettes with better perceptual
properties have been suggested in the literature
[@color:Harrower+Brewer:2003, @color:Zeileis+Hornik+Murrell:2009, @color:Smith+VanDerWalt:2015, @color:Crameri:2018]
and also made available in standard software packages including various
packages for \proglang{R} [@color:R] such as 
\pkg{RColorBrewer} [@color:RColorBrewer],
\pkg{rcartocolor} [@color:rcartocolor],
\pkg{viridis} [@color:viridis],
\pkg{wesanderson} [@color:wesanderson],
or \pkg{scico} [@color:scico] among many others.
In packages like \pkg{pals} [@color:pals] and \pkg{paletteer} [@color:paletteer]
many of these palettes are collected and provided with a unified interface.
Most of these palettes, however, are pre-existing palettes, stored as a
limited set of colors and interpolated as necessary. And even if specific
algorithms have been used in the first construction of the palettes, these
are often not reflected in the software implementations.

In contrast, the \pkg{colorspace} package [@color:colorspace] adopts a perceptually-based
approach -- the so-called HCL (hue--chroma--luminance) space -- and
establishes a broad range of color palettes through general and adaptable
trajectories in HCL space. In addition to constructing the palettes the HCL
space is also employed to visualize and assess the properties of a given
palette (beyond simple color swatches). The reason for choosing the HCL space
for all of these tools is that its three axes match those of the human visual system very well:
Hue (= type of color, dominant wavelength),
chroma (= colorfulness), luminance (= brightness), see Figure\ \ref{fig:hcl-properties}.

@@

To illustrate their usage two simple examples are shown using the qualitative `"Dark 3"`
and sequential `"Purples 3"` palettes that were also employed above
(see Figure\ \ref{fig:iris-diamonds}). For the first example, semi-transparent
shaded densities of the sepal length from the `iris` data are shown, grouped by species
(see the left panel of Figure\ \ref{fig:iris-diamonds}).

@@

And for the second example the sequential palette is used to code the cut levels
in a scatter of price by carat in the `diamonds` data (or rather a small subsample
thereof, see the right panel of Figure\ \ref{fig:iris-diamonds}). The scale function first
generates six colors but then drops the first color because the light gray is too light
here. (Alternatively, the chroma and luminance parameters could also be tweaked.)

@@

In Figure\ \ref{fig:allplots-sequential}, a heatmap (based on the well-known Maunga Whau volcano data) is used as a typical
application for a sequential palette. The elevation of the volcano is brought out clearly,
focusing with the dark colors on the higher elevations.
The other two displays show that hue is constant in the palette while luminance and chroma vary.
Luminance increases monotonically from dark to light (as required for a proper sequential palette).
Chroma is triangular-shaped which allows to better distinguish the middle colors in the palette
when compared to a monotonic chroma trajectory.

@@

For distinguishing colors in the center of the palette, two strategies can be employed:
(a) Hue can be varied as well by specifying an interval of hues in `h` (or beginning
hue `h1` and ending hue `h2`). More details are provided in the next section.
(b) Instead of a decreasing chroma a triangular chroma trajectory can be employed
from `c1` over `cmax` to `c2` (or equivalently a vector `c` of
length 3). This yields high-chroma colors in the middle of the palette that are
more easily distinguished from the dark and light extremes. See `"Blues 3"`, `"Purples 3"`,
`"Reds 3"`, and `"Greens 3"` in Figure\ \ref{fig:hcl-palettes-sequentials}.

@@

* _Single-hue:_ In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs.\ light gray).
* _Single-hue (advanced):_ The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
* _Multi-hue (advanced):_ As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

@@

* Hue is drawn in red and coordinates are indicated on the axis on the right
  with range $[0, 360]$ or (if necessary) $[-360, 360]$.
* Chroma is drawn in green with coordinates on the left axis. The range [0, 100]
  is used unless the palette necessitates higher chroma values.
* Luminance is drawn in blue with coordinates on the left axis in the range [0, 100].
